---
title: "Age-standardized rates"
author: "Connor Donegan"
date: "3/9/2022"
output: rmarkdown::html_vignette
header-includes:
   - \usepackage{amsmath}
vignette: >
  %\VignetteIndexEntry{Age-standardized rates}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: bib.bib
link-citations: yes
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  results = "hold", 
  collapse = TRUE, 
  eval = TRUE,
  fig.pos = 'h', 
  fig.align = 'center'
)
```

Age-standardization is used to improve comparison of disease incidence or mortality rates across populations. The term 'population' is used to reference any distinct grouping of people, whether it be geographic (e.g., two or more regions), temporal (e.g., two or more years), or social (e.g., two or more genders). Because disease and mortality risk vary strongly by age, we want to control for varying age distributions across groups. Examination of each age-specific rate is indispensable for any analysis, but age-standardized rates provide a single summary index of incidence or mortality risk. 

This vignette describes the method of direct age-standardization [@broemeling_2020] and then demonstrates implementation with **surveil**. The final section extends the methodology to the analysis of health inequality between two groups, where each group is age-stratified (e.g., two racial-ethnic groups in the same region).

## Data

The demonstration will use age-specific cancer incidence for the entire population of the United States of America, 1999-2017. 

```{r}
library(surveil)
data(cancer)
head(cancer)
```

We will also use the age distribution of the United States in the year 2000 (2000 U.S. standard million population, see `?standard`):

```{r}
data(standard)
print(standard)
```

Notice that the five-year age groups in the `cancer` data match the age groups provided by `standard`. 

In some cases, one is only interested in a subset of age groups. For example, imagine if our cancer data only included persons of 20-39 years. In thatWe see that relative inequality (AR) is about constant (as expected), the rate difference falls (because the total level of cancer risk has fallen over time), and excess cases increase over time (because the total population at risk is both growing and aging). case, we would also limit our standard population to the same age groups:

```{r}
standard_2 <- standard[6:9,]
print(standard_2)
```

Everything discussed below would proceed unchanged then, except we would replace `standard` with `standard_2`. 

## Direct age-standardization

### Methodology

Let $\theta_i$ be the disease risk in the $i^{th}$ age group, and let $\omega_i$ be the standard population count for that age group. Then the age-standardized risk is:
$$SR = \frac{\sum_i \theta_i \omega_i}{\sum_i \omega_i}$$
That is, age-standardization consists of multiplying actual age-specific risk levels by false, but fixed, population sizes.

### Crude age-standardized rates

To calculate crude age-standardized incidence rates, we use crude age-specific rates $r_i$ in place of the (modeled) risk levels $\theta_i$:
$$SR_{\text{crude}} = \frac{\sum_i r_i \omega_i}{\sum_i \omega_i}$$
We can calculate age-specific risks $r_i$ as follows: 
```{r}
standard$Age <- standard$age # make sure names match for the merge
cancer_2 <- merge(cancer, standard, by = "Age")  # properly join data together
cancer_2$crude_rate <- cancer_2$Count / cancer_2$Population  # r_i
cancer_2$crude_reweighted <- cancer_2$crude_rate * cancer_2$standard_pop  # r_i * w_i
```

We need the denominator (which equals one million in this case, by design):
```{r}
M <- sum(standard$standard_pop) # sum(w_i)
M
```

For 1999, the age-standardized cancer rate per 100,000 at-risk can be calculated as follows:

```{r}
c2_1999 <- cancer_2[grep(1999, cancer_2$Year),]
100e3 * sum(c2_1999$crude_reweighted) / M  # 100,000 * sum(r_i * w_i) / sum(w_i)
```

We can calculate crude age-standardized risk for each year, 1999-2017, as follows:

```{r}
crude_sr <- aggregate(crude_reweighted ~ Year, data = cancer_2, sum) 
crude_sr$SR <- 100e3 * crude_sr$crude_reweighted / M
print(crude_sr)
```

```{r fig.height = 5, fig.width = 6}
plot(crude_sr$Year, crude_sr$SR, type = "l", xlab = "", ylab = 'SR')
```

## Modeling age-specific risk

For research purposes, we almost always want to base our estimates of age-standardized risk on model-based inferences of age-specific risk $\theta_i$, as opposed to crude incidence rates. We model age-specific rates by providing `stan_rw` with the cancer data and telling it which column contains the time period indicator (`Year`) and which column contains the grouping variable (`Age`):

```{r}
fit <- stan_rw(cancer, time = Year, group = Age, 
               refresh = 0, iter = 1500)
```

The default plot method will return all of the age-specific cancer risk trends on the same plot; with many age groups and dissimilar risk levels, it is easier to understand the results if, instead, we use a grid of multiple small plots (`facet = TRUE`) and allow the scale of the y-axes to adjust for each age group (`facet_scales = "free"`):

```{r fig.width = 8, fig.height = 6}
plot(fit, 
     facet = TRUE,          # plot small multiples
     facet_scales = "free", # y-axes vary across plots
     base_size = 10,      # control text size 
     size = 0,            # removes crude rates from the plots
     scale = 100e3        # plot rates per 100,000
     )
```

The lines indicate estimates of risk $\theta_i$ (the means of the posterior probability distributions for $\theta_i$) and the shaded intervals correspond to 95\% credible intervals for $\theta_i$.

In addition to examining trends in age-specific risk (as above), we can also convert each age-specific trend to its annual percent change or cumulative percent change. This time, instead of using the default plotting style, we will use overlaid lines to depict the probability distributions for cumulative percent change:

```{r fig.width = 8, fig.height = 6}
fit_apc <- apc(fit)
plot(fit_apc, 
     style = "lines",
     base_size = 10,
     cum = TRUE)
```

Each line drawn on the figure is a draw (sample) from the joint posterior probability distribution for cumulative percent change in $\theta_i$: areas of high density of lines have higher probability than areas of low line density; likewise, the general spread of the lines at each time point indicates the range of plausible values.

## Age-standardizing model results

The `standardize` function takes a fitted model, plus the standard population, and returns standardized rates (SRs):

```{r}
fit_sr <- standardize(fit,
                        label = standard$age,
                        standard_pop = standard$standard_pop)
```

As usual, **surveil** provides a plotting method for the `fit_sr` object:

```{r fig.width = 5.5, fig.heigth = 4.5}
# load ggplot2 to enable additional plot customization
library(ggplot2)
plot(fit_sr, scale = 100e3, base_size = 10) + 
  labs(title = "US age-standardized cancer incidence per 100,000")
```

as well as a printing method:

```{r}
print(fit_sr)
```

To learn about the contents of `fit_sr`, see `?standardize` or explores its contents as you would a list (`names(fit_sr)`, `fit_sr$standard_summary`, `head(fit_sr$standard_samples`), etc.).

## Comparing age-stratified groups

### Measures of pairwise inequality

What if we are interested in measuring health inequality across two groups, each of which is age stratified? 

Let $\theta_{i,1}$ be the level of risk in the $i^{th}$ age group for the $1^{st}$ population, and $\theta_{i,2}$ be the risk for the $i^{th}$ age group in the $2^{nd}$ group; $P_{i,1}$ is the age-specific population at risk for group 1. Let group 1 be the group that is expected to have higher risk, and group 2 may be called the 'reference' group.

We can compare groups 1 and 2 on three measures of inequality:

 1. Standardized rate ratio: $RR = \frac{SR_1}{SR_2}$
 2. Standardized rate difference: $RD = SR_1 - SR_2$
 3. Total excess cases: $EC = \sum_i (\theta_{i,1} - \theta_{i,2}) * P_i = \sum_i RD_i * P_{i,1}$
 3. Total attributable risk: $AR = \frac{\sum_i (\theta_{i,1} - \theta_{i,2}) * P_i}{\sum_i \theta_{i,1} * P_{i,1}}$ 

The numerator in the formula for AR contains each age-specific rate difference multiplied by the age-specific population at risk for group 1. If we describe the rate difference as a measure of excess risk, then we can describe the (age-specific) quantity $RD_i * P_{i,1}$ as the number of "excess cases,", i.e., the excess risk expressed in terms of whole cases. The AR formula simply re-expresses excess risk as a fraction of total risk. This is convenient because the standardized RR is based on a fixed and false population distribution, whereas AR is based on the actual population distribution obtaining in reality, each year. Thus, we may often prefer AR over RR as a measure of relative inequality across two age-stratified groups. 

Pairwise measures of inequality have an important purpose (tracking health inequalities between historically or currently advantaged and disadvantaged groups), but we should keep in mind the limitations inherent in tracking the progress of any group by comparing that group to a moving target (particularly one that may move in an unfavorable direction). There are rarely only two groups of interest; to measure health inequality across multiple groups, consider Theil's index (see `?theil` and `vignette("demonstration")`).

### Demonstration

To demonstrate, we will use the same cancer data and create a fake comparison population. This will allow us to avoid introducing a new data set and it will give us control over the results.

Pretend that instead of having cancer incidence data for the entire USA, we have cancer data for two sub-populations, group A and group B. The original `cancer` data will serve for A (`cancer_a`), and we will create `cancer_b` so that its cancer incidence rate is, in relative terms, steadily above that of `cancer_a`. Group B will have smaller population than group A, and we will also add some extra 'noise' to the data.

```{r}
cancer_a <- cancer 
cancer_b <- cancer
## set the case count equal to 0.75 times that of group A
cancer_b$Count <- round(cancer_a$Count * 0.75) +  
  rpois(n = nrow(cancer), lambda = cancer_a$Count * 0.1) # adds a little noise to the data
## set the population at risk to 0.6 times that of group A
cancer_b$Population <- round(cancer_a$Population * 0.6) + 
  rpois(n = nrow(cancer), lambda = cancer_a$Population * 0.1)
```

Now we can model age-specific risk for Group B (`cancer_b`):

```{r}
fit_b <- stan_rw(cancer_b, time = Year, group = Age, 
                 refresh = 0, iter = 1500)
```

And then age-standardize the results:

```{r}
fit_sr_b <- standardize(fit_b,
                        label = standard$age,
                        standard_pop = standard$standard_pop)
```

Finally, we place the two age-standardized models into a list (preferably a named list), and pass the list to the `group_diff` function (see `?group_diff` for more details):

```{r}
fit_sr_list <- list(B = fit_sr_b, A = fit_sr)
ineq <- group_diff(fit_sr_list)
```

```{r fig.width = 7, fig.height = 2.5}
plot(ineq, base_size = 10)
```

We see that relative inequality (AR) is about constant (as expected), the rate difference falls (because the total level of cancer risk has fallen over time), and excess cases increase over time (because the total population at risk is both growing and aging).

```{r}
print(ineq)
```

## References



